Accessible, realistic genome simulation with selection using \stdpopsim

### 2025-03-19 analysis2_manuscript.tex
first edit
commas

minor edits

% PIs that were very involved in the project
\author[17]{Ilan Gronau}
\author[4]{Ryan N. Gutenkunst}
\author[11]{Jerome Kelleher}
\author[6]{Kirk E. Lohmueller}
\author[18]{Daniel R. Schrider}
\author[2,19]{Peter L. Ralph}
\author[2]{Andrew D. Kern}

\affil[*]{\small{These authors contributed equally to the paper.}}
\affil[1]{\small{Section for Molecular Ecology and Evolution, Globe Institute, University of Copenhagen, Denmark}}
\affil[2]{\small{Institute of Ecology and Evolution, University of Oregon, Eugene OR 97402, USA}}
\affil[3]{\small{Division of Genetics, Oregon National Primate Center, Oregon Health and Science University, Beaverton OR 97006, USA}}
\affil[4]{\small{Department of Molecular and Cellular Biology, University of Arizona, Tucson, AZ 85721, USA}}
\affil[5]{\small{Center for Systems Biology, New York University, New York NY, 10012 USA}}
\affil[6]{\small{Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles CA, USA}}
\affil[7]{\small{The Jackson Laboratory, Bar Harbor, ME, 04609 USA}}
\affil[8]{\small{Department of Computational Biology, Cornell University, Ithaca NY, USA}}
\affil[9]{\small{Department of Evolutionary Anthropology, University of Vienna, Vienna, Austria}}
\affil[10]{\small{Human Evolution and Archaeological Sciences (HEAS), University of Vienna, Vienna, Austria}}
\affil[11]{\small{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford OX3 7LF, UK}}
\affil[12]{\small{Instituto Nacional de Investigación Agropecuaria (INIA), Estación Experimental Las Brujas, Ruta 48 km 10, Canelones, Uruguay}}
\affil[13]{\small{Department of Genome Sciences, University of Washington, Seattle WA, USA}}
\affil[14]{\small{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, Edinburgh EH25 9RG, UK}}
\affil[15]{\small{Department of Biology, Pennsylvania State University, University Park, PA 16802, USA}}
\affil[16]{\small{Department of Integrative Biology, University of Wisconsin-Madison, Madison WI, USA}}
\affil[17]{\small{Efi Arazi School of Computer Science, Reichman University, Herzliya 4610101, Israel}}
\affil[18]{\small{Department of Genetics, University of North Carolina at Chapel Hill, Chapel Hill NC 27599, USA}}
\affil[19]{\small{Department of Data Science, University of Oregon, Eugene OR 97402, USA}}

@@

\section*{Abstract}
    \label{abstract}
    Natural selection is a fundamental evolutionary force that shapes patterns of genetic variation across species. 
    However, simulations incorporating realistic selection along heterogeneous genomes in complex demographic histories
    is challenging, limiting our ability to benchmark statistical methods aimed at detecting selection and to explore theoretical predictions.
    \stdpopsim is a community-maintained simulation library that already provides an extensive catalog of
    species-specific population genetic models.
    Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
    of natural selection, including background selection, selective sweeps,
    and arbitrary distributions of fitness effects (DFE) acting on annotated subsets of the genome (for instance, exons).
    This extension maintains \stdpopsim's core principles of reproducibility and accessibility while adding support
    for species-specific genomic annotations and published DFE estimates. 
    We demonstrate the utility of this framework by benchmarking methods for demographic inference,
    DFE estimation, and selective sweep detection across several species and scenarios. 
    Our results demonstrate the robustness of demographic inference methods to selection on linked sites, 
    reveal the sensitivity of DFE-inference methods to model assumptions, 
    and show how genomic features like recombination rate and the density of functional sequence influence power to detect selective sweeps.
    This extension to \stdpopsim{} provides a powerful, new resource for the population genetics community
    to explore the interplay between selection and other evolutionary forces in a reproducible, low-barrier framework.

\section*{Introduction}
    \label{introduction}
    % natural selection
    Natural selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of natural selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \citep[e.g.,][]{gillespie1991causes}.
    For instance, changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome 
    that can mimic the effects of selection \citep{simonsen1995properties},
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \citep{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn, selection can confound our ability to infer demographic 
    history from allele frequencies \citep{ewing2016consequences,schrider2016effects} and
    estimates of inverse coalescent rate \citep{schrider2016effects, johri2021impact, cousins2024accurate}.
    Thus, it is imperative to jointly account for the effects of selection
    and demography when inferring evolutionary history from genetic data \citep[e.g.,][]{sheehan2016deep,johri2020toward},
    yet tools for doing so remain limited.

    % simulation
    To meet the growing need for the interpretation,
    analysis, and exploration of realistic and complex evolutionary models,
    the field of population genetics has increasingly turned to simulation.
    Simulation in population genetics has a long history 
    including both backward-in-time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward-in-time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular, hernandez2008flexible, thornton2014c++, haller2019slim}. 
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{peischl2013accumulation, thornton2019polygenic, galloway2020few, zhang2020impact}, 
     to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical,hudson1994evidence,sabeti2002detecting},
    to explore the power and limitations of statistical methods \citep[e.g.,][]{przeworski2002signature},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods 
    \citep[e.g.,][]{beaumont2002approximate,lin2011distinguishing,kern2018diplos,mughal2019localizing,sanchez2021deep,wang2021automatic, zhang2023maladapt, smith2023dispersenn2}.
    However, joint simulation of complex demography and selection on realistic genomes is challenging,
    requiring many choices including the strength and rate of selection,
    and a parameterized demographic model,
    as well as implementation of genetic maps and other genomic features related to the strength of selection.
    This is a daunting task, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Furthermore, different simulation and inference tools may use different structures
    and parameter scalings for their models,
    making it difficult to directly compare across these tools and the studies that use them.

    % stdpopsim
    In light of the complexities associated with estimating and simulating
    population genetic models, it is no surprise that
    a lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community 
    \citep[e.g.,][]{speidel2021inferring, wang2021automatic, yang2022trustgwas, degiorgio2022spatially, browning2023fast, schweiger2023ultrafast, temple2024modeling, haag2025pandora}.
    Yet, previous releases of \stdpopsim did not include
    models of selection, which is a major limitation for more empirically motivated
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation \citep[e.g.][]{thornton2019polygenic,hartfield2024polygenic}, the effects of selective sweeps 
    \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, 
    and the impact of background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a simple, unified framework for simulating both neutral and
    non-neutral evolutionary processes and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Furthermore, the framework should include the complex realities of 
    genomes, including heterogeneous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Moreover, one should be able
    to model all of these features simultaneously using estimates and annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background,rodrigues2024shared}.
    %
    \stdpopsim{} provides a framework for such models,
    thus lowering the barrier to production of complex, multifaceted simulations,
    and allowing a wider range of researchers to make use of high-quality simulations
    in developing or testing their work.
    The project also holds as a central goal open, collaborative, and inclusive work,
    viewing capacity-building across the community as equal in importance to software development.

@@

    As with other aspects of \stdpopsim, we intend the selection models to be as biologically accurate as possible.
    Thus, the \stdpopsim catalog provides \texttt{Annotation} objects based on species' publicly available functional genomic elements 
    and \texttt{DFE} objects based on published DFE estimates (Figure \ref{fig:schematic}A).
    By using annotation objects, one can, for example, simulate populations where selected mutations (whose selection
    coefficients are drawn from a specified DFE) occur within annotation regions (e.g. exons), while all mutations
    occurring outside of these regions do not affect fitness.
    There are relatively few species with published DFE estimates in the literature.
    \stdpopsim currently has implemented DFEs for five species in the catalogue:
    \textit{Arabidopsis thaliana} \citep{huber2018gene},
    \textit{Drosophila melanogaster} \citep{ragsdale2016triallelic,huber2017determining,zhen2021greater},
    humans \citep{huber2017determining,kim2017inference,kyriazis2022using,castellano2019compmarison,castellano2019compmarison,rodrigues2024shared},
    \textit{Mus musculus} \citep{booker2021selective},
    and the vaquita porpoise \textit{Phocoena sinus} \citep{robinson2022critically}.
    While the DFEs are from a limited number of species, these DFEs can be applied to other species
    in the catalog if and when it is suitable. 
    For instance, one could simulate a butterfly species with a gamma-distributed DFE originally
    estimated from humans, or a human population with a DFE estimated from \textit{Drosophila}.
    Indeed, the DFE labeled \texttt{HomSap/Mixed\_K23} is recommended by \citet{kyriazis2022using}
    for ``general'' use across species (lacking more specific information).
    Furthermore, the user can specify a custom DFE and provide their own annotations
    of functional elements to simulate selection in a species for which we do not yet have 
    a published DFE included in the catalog.

@@

    % current numbers of DFEs / species / etc
    At this release (version 0.3.0) the \stdpopsim catalog comprises 27 species (i.e. genome representations)
    for which we have included 35 demographic models, 38 genetic maps, 12 DFEs, and 8 annotation objects. %ADK: checked this on 3/10/25
    Additions to the catalog are ongoing---we point the reader to our previously published 
    report detailing this effort \citep{lauterbur2023expanding}. We welcome contributions from the
    community and encourage those interested in contributing to visit https://popsim-consortium.github.io/stdpopsim-docs
    or contact one of the PopSim Consortium members for guidance.

    \section*{Inference of $N_e(t)$ in the context of selection}
    One of the most common applications of population genetic inference is to estimate
    the effective population size over time, $N_e(t)$, from genetic data. This can be done
    using a variety of methods, including the sequential Markovian coalescent
    \citep{li2011inference,Schiffels2020,terhorst2017robust}, 
    through use of the site frequency spectrum (SFS) \citep{liu2020stairway},
    as well as through identity by descent information \citep{santiago2020recent}.
    Since all of these methods assume neutrality, they are typically applied to genomic
    segments that are not expected to be directly affected by selection (e.g., by masking exons
    and conserved elements).
    However, natural selection acting on linked sites---even those that are masked out of the
    analysis---can bias estimates of $N_e(t)$
    away from census population sizes by increasing the 
    rate of coalescence in the genome \citep[e.g.][]{schrider2016effects}. 
    Using \stdpopsim, we can readily examine how different methods for inferring $N_e(t)$
    are influenced by selection on linked sites,
    which in the case of negative selection is commonly referred to as \emph{background selection}
    \citep{charlesworth1993effect,hudson1995deleterious}.

@@

    To do this we simulated human genomes with and without natural selection.
    In both scenarios, we ran three replicate simulations
    under the out of Africa (OOA) demographic model of
    \citet{ragsdale2019models} using a genetic map from the HapMap Project \citep{international2007second} (see \textbf{Methods}).
    Simulations with selection were implemented by modeling purifying selection on nonsynonymous mutations
    using a DFE inferred by \citet{kim2017inference}.
    A basic summary of nucleotide diversity from these simulations
    shows that the presence of selection decreases polymorphism only slightly
    at the genome-wide level, but more substantially within exonic regions (Figure \ref{fig:homsap-summary-stats}).
    %IG: no need to specify the two scenarios shown in fig
    %under both a constant population size and the out-of-Africa demographic model
    Following the simulation, inference of $N_e(t)$ was conducted from each of the six simulated datasets using four methods: 
    %IG: removed numbering and reordered to match fig 2
    \stairway \citep{liu2020stairway}, \msmc \citep{Schiffels2020}, \smcpp \citep{terhorst2017robust} and \gone \citep{santiago2020recent}.

    Although the estimates produced by the four methods differ from one another, 
    for each method the estimates produced from data with selection (Figure
    \ref{fig:human-demography}; orange) and without selection (Figure
    \ref{fig:human-demography}; blue) are fairly similar.
    Thus, this particular parameterization of background selection and demography,
    in combination with selection only at exonic sites, %in the out-of-Africa demographic model IG: it's probably not just this specific demo model
    does not appear to notably bias the estimates produced by these methods.
    This makes sense in light of our observation that genetic diversity is mostly affected
    within exonic regions and presumably sites very closely linked (Figure \ref{fig:homsap-summary-stats}),
    thus masking of exons should be sufficient to remove the major effects of selection. 
    As noted in previous studies that compared demographic inference methods
    on neutral simulations \citep{adrion2020community},
    we see that each method has its strengths and weaknesses.
    Methods based on the sequential Markovian coalescent (\smcpp and \msmc)
    appear to produce the most accurate trajectories of $N_e(t)$ overall
    in this setting, although they tend to under-estimate the population
    bottleneck associated with the population split.
    On the other hand, \stairway, which utilizes the site frequency spectrum,
    produces more noisy estimates,
    but it does not appear to under-estimate the population bottleneck.
    The inference of \gone, which utilizes identity by descent,
    is targeted toward demographic changes in more recent time frames ($\sim 200$ generations),
    explaining its noisy estimates for all but the most recent time periods.

@@

    A challenge when comparing different inferred and simulated distributions of the selection coefficient ($s$)
    is differing conventions in the literature regarding its definition.
    The simulator \slim and the inference tool \grapes define the selection coefficient $s$ such that a homozygote has fitness $1+s$.
    On the other hand, \dadicli and \polydfe define the selection coefficient such that a homozygote has fitness $1+ 2s$.
    Moreover, inferred distributions of selection coefficients are typically scaled by an inferred ancestral
    effective population size $N_e$, and different methods assume different scaling.
    \dadicli infers the distribution of $2 N_e s$ while \polydfe and \grapes infer the distribution of $4 N_e s$.
    To allow direct comparison, here we have normalized all inference results to the convention used by \slim
    (also used by \stdpopsim; see \textbf{Methods}).
    %IG - citations suggested in earlier versions to be added (I'm not sure they're necessary)
    % precursor of polyDFE  from 2017:     https://academic.oup.com/genetics/article/207/3/1103/5930676
    % fastDFE - the successor to polyDFE:  https://academic.oup.com/mbe/article/41/5/msae070/7641109
    % paper that uses GRAPES from 2020:    https://academic.oup.com/genetics/article/216/2/559/6066183

@@

     %sweeep power figure
    \begin{figure}[b!]
        \centering
        \includegraphics[width=0.8 \textwidth]{figures/sweeps/chr1_power.pdf}
        \caption{
        %IG: better not mention 100 if we're filtering out 7 segments
        %IG: removed citations from fig captions; they are prominently specified in main text
        Power to detect selective sweeps along human
        chromosome 1. Genomic segments were simulated with sweeps under a three population out-of-Africa model
        %\citep{gutenkunst2009inferring}
        and with background selection from deleterious mutations in exons.
        Three methods for detecting sweeps were applied to simulated data:
        \sweepfinder (top row -- labelled CLR),
        \diploshic (middle row),
        and reduced diversity ($\pi$) (bottom row).
        Power (true positive rate) is shown for these methods for the CEU and YRI
        samples (left and right respectively).
        The thresholds of the test statistics were set to control for
        $5\%$ false positive rate under a neutral null model (blue)
        and a null model with background selection from deleterious mutations in exons (red).
        Also shown are heatmaps of local recombination rate and exon density along the
        chromosome (bottom).
        }
        \label{fig:chr1_power}
    \end{figure}

@@

\section*{Funding}
    \label{funding}
    % we got too many people to do initials
    M.~Rodgrigues, S.~Tittes, N.~Pope, and A.~Kern were supported in part by NIH award R35GM148253.
    R.~Gutenkunst and L.~Tran were supported in part by NIH award R35GM149235.
    D.~Schrider was supported in part by NIH award R35GM138286.
    K.~Lohmueller was supported by NIH award R35GM119856.
    P.~Ralph, B.~Haller, A.~Kern, and B.~Jefferey
    were supported in part by NIH award R01HG012473.
    J.~Kelleher was supported by EPSRC grant EP/X024881/1
    and the Robertson Foundation.

>>>>

% PIs that were very involved in the project
\author[17]{Ilan Gronau}
\author[4]{Ryan N. Gutenkunst}
\author[11]{Jerome Kelleher}
\author[6]{Kirk E. Lohmueller}
\author[18]{Daniel R. Schrider}
\author[2,19]{Peter L. Ralph}
\author[2,$\heartsuit$]{Andrew D. Kern}

\affil[*]{\small{These authors contributed equally to the paper.}}
\affil[$\heartsuit$]{\small{To whom correspondence should be addressed.}} 
\affil[1]{\small{Section for Molecular Ecology and Evolution, Globe Institute, University of Copenhagen, Denmark}}
\affil[2]{\small{Institute of Ecology and Evolution, University of Oregon, Eugene, OR, 97402, USA}}
\affil[3]{\small{Division of Genetics, Oregon National Primate Center, Oregon Health and Science University, Beaverton, OR, 97006, USA}}
\affil[4]{\small{Department of Molecular and Cellular Biology, University of Arizona, Tucson, AZ, 85721, USA}}
\affil[5]{\small{Center for Genomics \& Systems Biology, New York University, New York, NY, 10003, USA}}
\affil[6]{\small{Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, CA, USA}}
\affil[7]{\small{The Jackson Laboratory, Bar Harbor, ME, 04609, USA}}
\affil[8]{\small{Department of Computational Biology, Cornell University, Ithaca, NY, USA}}
\affil[9]{\small{Department of Evolutionary Anthropology, University of Vienna, Vienna, Austria}}
\affil[10]{\small{Human Evolution and Archaeological Sciences (HEAS), University of Vienna, Vienna, Austria}}
\affil[11]{\small{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford OX3 7LF, UK}}
\affil[12]{\small{Instituto Nacional de Investigación Agropecuaria (INIA), Estación Experimental Las Brujas, Ruta 48 km 10, Canelones, Uruguay}}
\affil[13]{\small{Department of Genome Sciences, University of Washington, Seattle, WA, 98195, USA}}
\affil[14]{\small{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, Edinburgh EH25 9RG, UK}}
\affil[15]{\small{Department of Biology, Pennsylvania State University, University Park, PA, 16802, USA}}
\affil[16]{\small{Department of Integrative Biology, University of Wisconsin-Madison, Madison, WI, USA}}
\affil[17]{\small{Efi Arazi School of Computer Science, Reichman University, Herzliya 4610101, Israel}}
\affil[18]{\small{Department of Genetics, University of North Carolina at Chapel Hill, Chapel Hill, NC, 27599, USA}}
\affil[19]{\small{Department of Data Science, University of Oregon, Eugene, OR, 97402, USA}}

@@

\section*{Abstract}
    \label{abstract}
    Selection is a fundamental evolutionary force that shapes patterns of genetic variation across species.
    However, simulations incorporating realistic selection along heterogeneous genomes in complex demographic histories
    is challenging, limiting our ability to benchmark statistical methods aimed at detecting selection and to explore theoretical predictions.
    \stdpopsim is a community-maintained simulation library that already provides an extensive catalog of
    species-specific population genetic models.
    Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
    of selection, including background selection, selective sweeps,
    and arbitrary distributions of fitness effects (DFE) acting on annotated subsets of the genome (for instance, exons).
    This extension maintains \stdpopsim's core principles of reproducibility and accessibility while adding support
    for species-specific genomic annotations and published DFE estimates.
    We demonstrate the utility of this framework by benchmarking methods for demographic inference,
    DFE estimation, and selective sweep detection across several species and scenarios.
    Our results demonstrate the robustness of demographic inference methods to selection on linked sites,
    reveal the sensitivity of DFE-inference methods to model assumptions,
    and show how genomic features, like recombination rate and the density of functional sequence, influence power to detect selective sweeps.
    This extension to \stdpopsim{} provides a powerful, new resource for the population genetics community
    to explore the interplay between selection and other evolutionary forces in a reproducible, low-barrier framework.

\section*{Introduction}
    \label{introduction}
    % selection
    Selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \citep[e.g.,][]{gillespie1991causes}.
    For instance, changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome
    that can mimic the effects of selection \citep{simonsen1995properties},
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \citep{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn, selection can confound our ability to infer demographic
    history from allele frequencies \citep{ewing2016consequences,schrider2016effects} and
    estimates of inverse coalescent rate \citep{schrider2016effects, johri2021impact, cousins2024accurate}.
    Thus, it is imperative to jointly account for the effects of selection
    and demography when inferring evolutionary history from genetic data \citep[e.g.,][]{sheehan2016deep,johri2020toward},
    yet the number of tools for doing so remain limited.

    % simulation
    To meet the growing need for the interpretation,
    analysis, and exploration of realistic and complex evolutionary models,
    the field of population genetics has increasingly turned to simulation.
    Simulation in population genetics has a long history
    including both backward-in-time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward-in-time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular, hernandez2008flexible, thornton2014c++, haller2019slim, gaynor2021AlphaSimR}.
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{peischl2013accumulation, thornton2019polygenic, galloway2020few, zhang2020impact},
     to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical,hudson1994evidence,sabeti2002detecting},
    to explore the power and limitations of statistical methods \citep[e.g.,][]{przeworski2002signature, macleod2014effects},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods
    \citep[e.g.,][]{beaumont2002approximate,lin2011distinguishing,kern2018diplos,mughal2019localizing,sanchez2021deep,wang2021automatic, zhang2023maladapt, smith2023dispersenn2}.
    However, joint simulation of complex demography and selection on realistic genomes is challenging,
    requiring many choices including the strength and rate of selection,
    and a parameterized demographic model,
    as well as implementation of genetic maps and other genomic features related to the strength of selection.
    This is a daunting task, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Furthermore, different simulation and inference tools may use different structures
    and parameter scalings for their models,
    making it difficult to directly compare across these tools and the studies that use them.

    % stdpopsim
    In light of the complexities associated with estimating and simulating
    population genetic models, it is no surprise that
    a lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has been widely adopted by the community
    \citep[e.g.,][]{speidel2021inferring, wang2021automatic, yang2022trustgwas, degiorgio2022spatially, browning2023fast, schweiger2023ultrafast, temple2024modeling, haag2025pandora}.
    Yet, previous releases of \stdpopsim did not include
    models of selection, which is a major limitation for more empirically motivated
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation \citep[e.g.][]{thornton2019polygenic,hartfield2024polygenic}, the effects of selective sweeps
    \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft},
    and the impact of background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a simple, unified framework for simulating both neutral and
    non-neutral evolutionary processes and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible.
    Furthermore, the framework should include the complex realities of
    genomes, including heterogeneous recombination rates,
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Moreover, one should be able
    to model all of these features simultaneously using estimates and annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background,rodrigues2024shared}.
    %
    \stdpopsim{} provides a framework for such models,
    thus lowering the barrier to production of complex, multifaceted simulations,
    and allowing a wider range of researchers to make use of high-quality simulations
    in developing or testing their work.
    The project also holds as a central goal open, collaborative, and inclusive work,
    viewing capacity-building across the community as equal in importance to software development.

@@

    As with other aspects of \stdpopsim, we intend the selection models to be as biologically accurate as possible.
    Thus, the \stdpopsim catalog provides \texttt{Annotation} objects based on species' publicly available functional genomic elements
    and \texttt{DFE} objects based on published DFE estimates (Figure \ref{fig:schematic}A).
    By using annotation objects, one can, for example, simulate populations where selected mutations (whose selection
    coefficients are drawn from a specified DFE) occur within annotation regions (e.g. exons), while all mutations
    occurring outside of these regions do not affect fitness.
    There are relatively few species with published DFE estimates in the literature.
    \stdpopsim currently has implemented DFEs for five species in the catalog:
    \textit{Arabidopsis thaliana} \citep{huber2018gene},
    \textit{Drosophila melanogaster} \citep{ragsdale2016triallelic,huber2017determining,zhen2021greater},
    humans \citep{huber2017determining,kim2017inference,kyriazis2022using,castellano2019compmarison,castellano2019compmarison,rodrigues2024shared},
    \textit{Mus musculus} \citep{booker2021selective},
    and the vaquita porpoise \textit{Phocoena sinus} \citep{robinson2022critically}.
    While the DFEs are from a limited number of species, these DFEs can be applied to other species
    in the catalog if and when it is suitable.
    For instance, one could simulate a butterfly species with a gamma-distributed DFE originally
    estimated from humans, or a human population with a DFE estimated from \textit{Drosophila}.
    Indeed, the DFE labeled \texttt{HomSap/Mixed\_K23} is recommended by \citet{kyriazis2022using}
    for ``general'' use across species (lacking more specific information).
    Furthermore, the user can specify a custom DFE and provide their own annotations
    of functional elements to simulate selection in a species for which we do not yet have
    a published DFE included in the catalog.

@@

    % current numbers of DFEs / species / etc
    At this release (version 0.3.0) the \stdpopsim catalog comprises 27 species (i.e. genome representations)
    for which we have included 35 demographic models, 38 genetic maps, 12 DFEs, and 8 annotation objects. %ADK: checked this on 3/10/25
    Additions to the catalog are ongoing---we point the reader to our previously published
    report detailing this effort \citep{lauterbur2023expanding}. We welcome contributions from the
    community and encourage those interested in contributing to visit \url{https://popsim-consortium.github.io/stdpopsim-docs}
    or contact one of the PopSim Consortium members for guidance.

    \section*{Inference of $N_e(t)$ in the context of selection}
    One of the most common applications of population genetic inference is to estimate
    the effective population size over time, $N_e(t)$, from genetic data. This can be done
    using a variety of methods, including the sequential Markovian coalescent
    \citep{li2011inference,Schiffels2020,terhorst2017robust},
    through use of the site frequency spectrum (SFS) \citep{liu2020stairway},
    as well as through identity by descent information \citep{santiago2020recent}.
    Since all of these methods assume neutrality, they are typically applied to genomic
    segments that are not expected to be directly affected by selection (e.g., by masking exons
    and conserved elements).
    However, selection acting on linked sites---even those that are masked out of the
    analysis---can bias estimates of $N_e(t)$
    away from census population sizes by increasing the
    rate of coalescence in the genome \citep[e.g.][]{schrider2016effects}.
    Using \stdpopsim, we can readily examine how different methods for inferring $N_e(t)$
    are influenced by selection on linked sites,
    which in the case of negative selection is commonly referred to as \emph{background selection}
    \citep{charlesworth1993effect,hudson1995deleterious}.

@@

    To do this we simulated human genomes with and without selection.
    In both scenarios, we ran three replicate simulations
    under the out of Africa (OOA) demographic model of
    \citet{ragsdale2019models} using a genetic map from the HapMap Project \citep{international2007second} (see \textbf{Methods}).
    Simulations with selection were implemented by modeling purifying selection on nonsynonymous mutations
    using a DFE inferred by \citet{kim2017inference}.
    A basic summary of nucleotide diversity from these simulations
    shows that the presence of selection decreases polymorphism only slightly
    at the genome-wide level, but more substantially within exonic regions (Figure \ref{fig:homsap-summary-stats}).
    %IG: no need to specify the two scenarios shown in fig
    %under both a constant population size and the out-of-Africa demographic model
    Following the simulation, inference of $N_e(t)$ was conducted from each of the six simulated datasets using four methods:
    %IG: removed numbering and reordered to match fig 2
    \stairway \citep{liu2020stairway}, \msmc \citep{Schiffels2020}, \smcpp \citep{terhorst2017robust} and \gone \citep{santiago2020recent}.

    Although the estimates produced by the four methods differ from one another,
    for each method the estimates produced from data with selection (Figure
    \ref{fig:human-demography}; orange) and without selection (Figure
    \ref{fig:human-demography}; blue) are fairly similar.
    Thus, this particular parameterization of background selection and demography,
    in combination with selection only at exonic sites, %in the out-of-Africa demographic model IG: it's probably not just this specific demo model
    does not appear to notably bias the estimates produced by these methods.
    This makes sense in light of our observation that genetic diversity is mostly affected
    within exonic regions and presumably sites very closely linked (Figure \ref{fig:homsap-summary-stats}),
    thus masking of exons should be sufficient to remove the major effects of selection.
    As noted in previous studies that compared demographic inference methods
    on neutral simulations \citep{adrion2020community},
    we see that each method has its strengths and weaknesses.
    Methods based on the sequential Markovian coalescent (\smcpp and \msmc)
    appear to produce the most accurate trajectories of $N_e(t)$ overall
    in this setting, although they tend to under-estimate the population
    bottleneck associated with the population split.
    On the other hand, \stairway, which utilizes the site frequency spectrum,
    produces more noisy estimates,
    but it does not appear to under-estimate the population bottleneck.
    The inference of \gone, which utilizes linkage-disequilibrium,
    is targeted toward demographic changes in more recent time frames ($\sim 200$ generations),
    explaining its noisy estimates for all but the most recent time periods.

@@

    A challenge when comparing different inferred and simulated distributions of the selection coefficient ($s$)
    is differing conventions in the literature regarding its definition.
    The simulator \slim and the inference tool \grapes define the selection coefficient $s$ such that a homozygote has fitness $1+s$.
    On the other hand, \dadicli and \polydfe define the selection coefficient such that a homozygote has fitness $1+2s$.
    Moreover, inferred distributions of selection coefficients are typically scaled by an inferred ancestral
    effective population size $N_e$, and different methods assume different scaling.
    \dadicli infers the distribution of $2 N_e s$ while \polydfe and \grapes infer the distribution of $4 N_e s$.
    To allow direct comparison, here we have normalized all inference results to the convention used by \slim
    (also used by \stdpopsim; see \textbf{Methods}).
    %IG - citations suggested in earlier versions to be added (I'm not sure they're necessary)
    % precursor of polyDFE  from 2017:     https://academic.oup.com/genetics/article/207/3/1103/5930676
    % fastDFE - the successor to polyDFE:  https://academic.oup.com/mbe/article/41/5/msae070/7641109
    % paper that uses GRAPES from 2020:    https://academic.oup.com/genetics/article/216/2/559/6066183

@@

     %sweeep power figure
    \begin{figure}[b!]
        \centering
        \includegraphics[width=0.8 \textwidth]{figures/sweeps/chr1_power.pdf}
        \caption{
        %IG: better not mention 100 if we're filtering out 7 segments
        %IG: removed citations from fig captions; they are prominently specified in main text
        Power to detect selective sweeps along human
        chromosome 1. Genomic segments were simulated with sweeps under a three population out-of-Africa model
        %\citep{gutenkunst2009inferring}
        and with background selection from deleterious mutations in exons.
        Three methods for detecting sweeps were applied to simulated data:
        \sweepfinder (top row -- labeled CLR),
        \diploshic (middle row),
        and reduced diversity ($\pi$) (bottom row).
        Power (true positive rate) is shown for these methods for the CEU and YRI
        samples (left and right respectively).
        The thresholds of the test statistics were set to control for
        $5\%$ false positive rate under a neutral null model (blue)
        and a null model with background selection from deleterious mutations in exons (red).
        Also shown are heatmaps of local recombination rate and exon density along the
        chromosome (bottom).
        }
        \label{fig:chr1_power}
    \end{figure}

@@

\section*{Funding}
    \label{funding}
    % we got too many people to do initials
    M.~Rodgrigues, S.~Tittes, N.~Pope, and A.~Kern were supported in part by NIH award R35GM148253.
    R.~Gutenkunst and L.~Tran were supported in part by NIH award R35GM149235.
    D.~Schrider was supported in part by NIH award R35GM138286.
    K.~Lohmueller was supported by NIH award R35GM119856.
    P.~Ralph, B.~Haller, A.~Kern, and B.~Jefferey
    were supported in part by NIH award R01HG012473.
    J.~Kelleher was supported by EPSRC grant EP/X024881/1
    and the Robertson Foundation.
    G.~Gorjanc and Y.~Zhang were supported in part by BBSRC (BBS/E/RL/230001A, BBS/E/RL/230001C, BB/T014067/1),
    Chinese Scolarship Council (CSC) and Genus.
